---
title: "Predicting BMW Stock Price Direction Using Machine Learning"
author: "Xinyi Yang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  fig.align  = "center"
)
```

```{r libraries}
pkgs <- c("tidyverse", "tidymodels", "discrim", "kknn",
          "vip", "patchwork", "lubridate", "ranger",
          "corrplot", "glmnet", "MASS")
install.packages(setdiff(pkgs, rownames(installed.packages())))

library(tidyverse)
library(tidymodels)
library(discrim)
library(kknn)
library(vip)
library(patchwork)
library(lubridate)
library(ranger)
library(glmnet)
library(corrplot)

tidymodels_prefer()
set.seed(131)
```

---

# Introduction

## Background and Motivation

Bayerische Motoren Werke AG (BMW) is one of the world's leading premium
automobile and motorcycle manufacturers. Its stock (ticker: BMW.DE) has been
traded on the Frankfurt Stock Exchange since the 1960s. Predicting the
short-term direction of stock prices is a classic and challenging problem in
quantitative finance — one that has attracted enormous interest from both
academic researchers and industry practitioners.

This project asks a focused, practically relevant question:

> **Can we predict whether BMW's stock price will rise (go up) or fall (go
> down) on the next trading day, using only information available at the
> close of the current trading day?**

This is framed as a binary classification problem. The outcome variable
`Target` equals 1 if the next-day closing price is higher than the current
day's closing price, and 0 otherwise.

## Dataset Description

The dataset is the *BMW Stock Advanced Features* dataset obtained from
Kaggle. After removing any incomplete rows, 7,390 trading-day observations
remain, each representing one trading day for BMW AG stock (ticker: BMW.DE)
on the Frankfurt Stock Exchange.

**Formal citation:**

> Sumon, Md Mahfuz. (2024). *BMW Car with Advanced Features* [Data set].
> Kaggle. <https://www.kaggle.com/datasets/mdmahfuzsumon/bmw-car-with-advanced-features>

Each row contains 31 features grouped as follows:

| Group | Variables |
|---|---|
| **Price** | `Open`, `High`, `Low`, `Close`, `Adj Close` |
| **Returns** | `Return`, `Log_Return` |
| **Momentum** | `Momentum_7`, `Momentum_30` |
| **Moving averages** | `MA7`, `MA30`, `MA100`, `EMA20`, `EMA50` |
| **Volatility** | `Volatility_7`, `Volatility_30` |
| **Spread / Range** | `High_Low_Spread`, `Open_Close_Spread`, `Price_Range_Pct` |
| **Lagged prices** | `Lag1`, `Lag2`, `Lag3` |
| **Time** | `Year`, `Month`, `Day`, `DayOfWeek` |
| **Volume** | `Volume`, `Volume_Change`, `Volume_MA7` |
| **Outcome** | `Target` (1 = next day up, 0 = next day down) |

## Why Classification? Why This Metric?

Stock return prediction is notoriously difficult because financial markets are
close to efficient: publicly available information is rapidly priced in,
leaving little systematic predictability. Rather than trying to predict the
exact magnitude of tomorrow's return (regression), we focus on its
direction (classification), which is a simpler and more actionable goal.

Following the instructor's recommendation to adopt a more informative
evaluation metric, we use ROC-AUC (area under the receiver operating
characteristic curve) as the primary performance metric throughout this
project. ROC-AUC is preferred over accuracy because:

1. It is **threshold-independent**: it measures discrimination ability
   across all possible decision thresholds, rather than fixing a cutoff at
   0.5.
2. It is **robust to class imbalance**: even though our classes are nearly
   balanced (50.8% down, 49.2% up), accuracy can be misleading when one
   class is easier to predict than the other.
3. It has a **natural probabilistic interpretation**: ROC-AUC equals the
   probability that the model ranks a randomly chosen positive example
   higher than a randomly chosen negative example.

An ROC-AUC of 0.5 corresponds to random guessing; 1.0 is a perfect
classifier.

---

# Exploratory Data Analysis

```{r load-data}
bmw <- read_csv("bmw_advanced_features.csv") %>%
  mutate(
    Target     = factor(Target, levels = c("1", "0"),
                        labels = c("Up", "Down")),
    Date       = as.Date(Date),
    DayOfWeek  = factor(DayOfWeek,
                        labels = c("Mon","Tue","Wed","Thu","Fri"))
  )

cat("Observations:", nrow(bmw), "\n")
cat("Features:    ", ncol(bmw), "\n")
cat("Date range:  ", format(min(bmw$Date)), "to", format(max(bmw$Date)), "\n")
```

## Missing Data

```{r missing-data}
# Count missing values per column
missing_summary <- bmw %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to = "Variable", values_to = "Missing") %>%
  filter(Missing > 0) %>%
  mutate(Pct = round(Missing / nrow(bmw) * 100, 2)) %>%
  arrange(desc(Missing))

if (nrow(missing_summary) == 0) {
  cat("No missing values detected in any column.\n")
} else {
  knitr::kable(missing_summary,
               caption = "Table 0. Variables with missing values")
}

# Check for Inf values
inf_cols <- bmw %>%
  summarise(across(where(is.numeric), ~ sum(is.infinite(.)))) %>%
  pivot_longer(everything(),
               names_to = "Variable", values_to = "Inf_Count") %>%
  filter(Inf_Count > 0)

if (nrow(inf_cols) > 0) {
  cat("\nColumns with Inf/-Inf values:\n")
  print(inf_cols)
}
```

The dataset contains no traditional missing (NA) values: all 7,390
rows are complete. However, the `Volume_Change` column contains
`Inf` and `-Inf` values on days when the previous day's trading volume
was zero (division by zero). These infinite values cannot be normalised or
used by any of our models. We therefore remove `Volume_Change` entirely
from the modelling pipeline before fitting any model. The information it
was meant to capture (relative volume change) is partially represented by
`Volume_MA7`, which remains in the predictor set.

The moving average and volatility features (`MA7`, `MA30`, `Volatility_7`,
etc.) require a warm-up period of 7–30 trading days before they can be
computed. Any resulting NAs in the first few rows are handled by
`step_impute_median()` in the recipe.



```{r fig-target, fig.cap="Figure 1. Class balance of the Target variable."}
bmw %>%
  count(Target) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = Target, y = n, fill = Target)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%d\n(%.1f%%)", n, pct)),
            vjust = -0.4, size = 4.5) +
  scale_fill_manual(values = c("Up" = "#2166ac", "Down" = "#d73027")) +
  scale_y_continuous(limits = c(0, 4200)) +
  labs(
    title = "Distribution of Target Variable",
    subtitle = "1 = Next-day price Up | 0 = Next-day price Down",
    x = "Direction",
    y = "Count"
  ) +
  theme_minimal(base_size = 13)
```

The dataset is remarkably balanced: 3,628 "Up" days (49.1%) vs. 3,762
"Down" days (50.9%). This near-perfect balance means that a naive classifier
that always predicts "Down" achieves only 50.9% accuracy. It also means that class-imbalance corrections are not required.

## BMW Closing Price Over Time

```{r fig-price, fig.cap="Figure 2. BMW adjusted closing price 1997–2026."}
bmw %>%
  ggplot(aes(x = Date, y = `Adj Close`)) +
  geom_line(color = "#2166ac", alpha = 0.8, linewidth = 0.4) +
  geom_smooth(method = "loess", span = 0.15,
              color = "#d73027", se = FALSE, linewidth = 1) +
  labs(
    title   = "BMW Adjusted Closing Price (1997–2026)",
    x       = "Date",
    y       = "Adjusted Close (EUR)"
  ) +
  theme_minimal(base_size = 13) +
  annotate("rect", xmin = as.Date("2008-01-01"),
           xmax = as.Date("2009-06-01"),
           ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "red") +
  annotate("text", x = as.Date("2008-09-01"), y = 100,
           label = "GFC\n2008–09", size = 3.2, color = "red")
```

BMW's stock grew from roughly EUR 11 (adjusted) in 1997 to a peak of about
EUR 122 in 2021, before a partial retreat. The red-shaded band marks the
2008–2009 Global Financial Crisis, during which the stock lost approximately
60% of its value. The LOESS trend line (red) confirms the long-term upward
trajectory punctuated by cyclical downturns.

## Daily Return Distribution by Target Class

```{r fig-return, fig.cap="Figure 3. Daily return distributions for Up vs Down days."}
p1 <- bmw %>%
  ggplot(aes(x = Return, fill = Target)) +
  geom_density(alpha = 0.55) +
  scale_fill_manual(values = c("Up" = "#2166ac", "Down" = "#d73027")) +
  labs(title = "Daily Return Distribution",
       x = "Return", y = "Density", fill = "Next Day") +
  theme_minimal(base_size = 12)

p2 <- bmw %>%
  ggplot(aes(x = Target, y = Return, fill = Target)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c("Up" = "#2166ac", "Down" = "#d73027")) +
  labs(title = "Return by Next-Day Direction",
       x = "Next Day", y = "Return") +
  theme_minimal(base_size = 12)

p1 + p2
```

Both distributions are centred near zero and nearly symmetrical, consistent with the efficient market hypothesis. However, "Up" days exhibit a slightly higher mean return (0.063%) vs. "Down" days (0.013%), suggesting a weak positive momentum signal. The market tends to continue rising on days following small gains.

## Volatility Patterns Over Time

```{r fig-vol, fig.cap="Figure 4. 30-day rolling volatility over time, coloured by Target."}
bmw %>%
  ggplot(aes(x = Date, y = Volatility_30, color = Target)) +
  geom_line(alpha = 0.5, linewidth = 0.3) +
  scale_color_manual(values = c("Up" = "#2166ac", "Down" = "#d73027")) +
  labs(
    title  = "30-Day Rolling Volatility (1997–2026)",
    x      = "Date",
    y      = "Volatility (std of 30-day returns)",
    color  = "Next Day"
  ) +
  theme_minimal(base_size = 13)
```

Volatility clusters are clearly visible around major market events: the
dot-com crash (2000–2002), the GFC (2008–2009), and COVID-19 (2020).
During high-volatility regimes, both "Up" and "Down" labels are densely
interleaved, confirming that predicting direction is harder in turbulent
markets.

## Correlation Matrix of Continuous Predictors

```{r fig-corr, fig.cap="Figure 5. Lower-triangle Pearson correlation matrix.", fig.height=7}
bmw %>%
  select(Return, Log_Return, Momentum_7, Momentum_30,
         Volatility_7, Volatility_30, High_Low_Spread,
         Open_Close_Spread, Lag1, Lag2, Lag3) %>%
  cor(use = "complete.obs") %>%
  corrplot(
    method = "circle",
    type   = "lower",
    diag   = FALSE,
    tl.cex = 0.85,
    tl.col = "black",
    addCoef.col = "black",
    number.cex  = 0.65,
    title  = "Correlation Matrix of Numeric Features",
    mar    = c(0, 0, 1.5, 0)
  )
```

Key observations from the correlation matrix:

- `Return` and `Log_Return` are almost perfectly correlated ($r \approx 1$)
  — only one is needed as a predictor.
- `Volatility_7` and `Volatility_30` are moderately positively correlated
  ($r \approx 0.7$): short-term and long-term volatility regimes tend to
  co-move.
- `Momentum_7` and `Momentum_30` are weakly correlated ($r \approx 0.3$):
  they capture different time-scale trends.
- Lagged prices (`Lag1`, `Lag2`, `Lag3`) are highly correlated with each
  other (near-unit autocorrelation), which is expected for consecutive
  prices.
- None of the features has a strong linear correlation with the outcome,
  consistent with the near-random-walk nature of stock returns.

---

# Data Splitting and Cross-Validation

## Why Time-Aware Splitting Matters

Stock price data is a time series: observations are ordered
chronologically and are not exchangeable. Using standard random
cross-validation would cause data leakage — future information would
leak into the training set, artificially inflating performance estimates.

For example, if the model were trained on data from 2020 and validated on
data from 2010, it would implicitly "know" about market events that had not
yet occurred in the validation period.

To avoid this, we use two strategies:

1. **Temporal train/test split**: The earliest 80% of trading days (through
   approximately mid-2022) form the training set; the most recent 20%
   (mid-2022 to March 2026) form the held-out test set.

2. **Time-series cross-validation** (rolling-origin): Within the training
   set, we use `rsample::rolling_origin()` with a fixed training window and
   a 63-day (≈ 3-month) assessment window, stepping forward by 21 days
   (≈ 1 month) at a time. This produces approximately 20 CV folds, each
   respecting temporal order.

```{r split}
# Temporal split: first 80% train, last 20% test
n <- nrow(bmw)
cutoff <- floor(0.80 * n)

bmw_train <- bmw[1:cutoff, ]
bmw_test  <- bmw[(cutoff + 1):n, ]

cat("Training observations:", nrow(bmw_train),
    "(", format(min(bmw_train$Date)), "to",
    format(max(bmw_train$Date)), ")\n")
cat("Testing  observations:", nrow(bmw_test),
    "(", format(min(bmw_test$Date)), "to",
    format(max(bmw_test$Date)), ")\n")

# Rolling-origin CV: 5 folds only (fast but still time-aware)
bmw_folds <- rolling_origin(
  bmw_train,
  initial    = 4000,   # large fixed training window
  assess     = 252,    # ~1 year assessment window
  skip       = 300,    # step forward ~14 months between folds
  cumulative = FALSE
)

cat("Number of CV folds:", nrow(bmw_folds), "\n")
```

The training set covers roughly 25 years of data (1997–2022); the test set
covers the most recent 4 years (2022–2026). This split ensures that all
model selection and hyperparameter tuning decisions are made without ever
seeing the test data.

**Cross-validation**: We use 5 rolling-origin folds, each with a fixed
4,000-day (~16 years) training window and a 252-day (~1 year) assessment
window, stepping forward ~14 months between folds. This keeps computation
tractable while still respecting temporal order.

**A note on stratification**: For cross-sectional data (e.g., the Titanic
dataset), it is standard practice to stratify splits on the outcome variable
to preserve class proportions. However, for **time-series data**, stratified
random sampling would break the temporal ordering of observations, causing
data leakage (future data appearing in training folds). Since our dataset
is a time series, we use **temporal ordering** instead of stratification.
The outcome classes are nearly balanced (49.1% vs. 50.9%), so the risk of
class imbalance across folds is minimal.

---

# Model Fitting

## Recipe (Pre-processing)

```{r recipe}
# Drop Volume_Change from both sets before recipe
bmw_train <- bmw_train %>% select(-Volume_Change)
bmw_test  <- bmw_test  %>% select(-Volume_Change)

bmw_recipe <- recipe(Target ~ Return + Log_Return +
                       Momentum_7 + Momentum_30 +
                       Volatility_7 + Volatility_30 +
                       High_Low_Spread + Open_Close_Spread +
                       Price_Range_Pct +
                       Lag1 + Lag2 + Lag3 +
                       Volume_MA7 +
                       DayOfWeek + Month,
                     data = bmw_train) %>%
  step_rm(Log_Return) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())
```

**Design choices:**

- `Date`, `Open`, `High`, `Low`, `Close`, `Adj Close`, `Volume` (raw) are
  excluded: they either have no predictive content for direction, or their
  information is already captured by engineered features.
- `Log_Return` is removed because it is a near-perfect linear transformation
  of `Return` (correlation $\approx 1$).
- All predictors are centred and scaled (`step_normalize`) to ensure fair
  treatment across distance-based models (KNN) and to improve
  gradient-based optimisation (logistic regression).

## Models Fitted

We fit four classification models and compare them on ROC-AUC:

| Model | Key hyperparameter(s) | Rationale |
|---|---|---|
| Logistic Regression | Penalty $\lambda$ (elastic net) | Interpretable linear baseline |
| Linear Discriminant Analysis | — | Generative, assumes Gaussian features |
| Quadratic Discriminant Analysis | — | More flexible than LDA |
| Random Forest | `mtry`, `trees`, `min_n` | Handles non-linearity and interactions |

### Logistic Regression (Elastic Net)

```{r logistic}
log_spec <- logistic_reg(
  penalty = tune(),
  mixture = 0.5
) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

log_wf <- workflow() %>%
  add_recipe(bmw_recipe) %>%
  add_model(log_spec)

log_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 15)

log_res <- tune_grid(
  log_wf,
  resamples = bmw_folds,
  grid      = log_grid,
  metrics   = metric_set(roc_auc)
)

best_log <- select_best(log_res, metric = "roc_auc")
log_final <- finalize_workflow(log_wf, best_log)
log_fit   <- fit(log_final, data = bmw_train)
```

The tuning results below show how ROC-AUC varies across the penalty
grid. The best penalty is selected at the peak of the curve.

```{r log-tune-plot, fig.cap="Figure A. Logistic regression tuning: ROC-AUC vs. penalty."}
autoplot(log_res) +
  labs(title = "Logistic Regression Tuning Results",
       x = "Regularisation Penalty (log scale)",
       y = "ROC-AUC (CV)") +
  theme_minimal(base_size = 12)
```

### Linear Discriminant Analysis

```{r lda}
lda_spec <- discrim_linear() %>%
  set_engine("MASS") %>%
  set_mode("classification")

lda_wf  <- workflow() %>%
  add_recipe(bmw_recipe) %>%
  add_model(lda_spec)

lda_fit <- fit(lda_wf, data = bmw_train)
```

### Quadratic Discriminant Analysis

```{r qda}
qda_spec <- discrim_quad() %>%
  set_engine("MASS") %>%
  set_mode("classification")

qda_wf  <- workflow() %>%
  add_recipe(bmw_recipe) %>%
  add_model(qda_spec)

qda_fit <- fit(qda_wf, data = bmw_train)
```

### Random Forest (Tuned)

```{r rf, cache=TRUE}
rf_spec <- rand_forest(
  mtry  = tune(),
  trees = 200,
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_recipe(bmw_recipe) %>%
  add_model(rf_spec)

rf_grid <- grid_regular(
  mtry(range  = c(3, 8)),
  min_n(range = c(10, 30)),
  levels = 2
)

rf_res <- tune_grid(
  rf_wf,
  resamples = bmw_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc)
)

best_rf <- select_best(rf_res, metric = "roc_auc")
rf_final <- finalize_workflow(rf_wf, best_rf)
rf_fit   <- fit(rf_final, data = bmw_train)
```

```{r rf-tune-plot, fig.cap="Figure B. Random Forest tuning: ROC-AUC vs. mtry and min_n."}
autoplot(rf_res) +
  labs(title = "Random Forest Tuning Results",
       y = "ROC-AUC (CV)") +
  theme_minimal(base_size = 12)
```

---

# Model Selection and Performance

## Cross-Validation ROC-AUC (Training)

```{r cv-results}
log_cv <- show_best(log_res, metric = "roc_auc", n = 1) %>%
  mutate(Model = "Logistic Regression")
rf_cv  <- show_best(rf_res,  metric = "roc_auc", n = 1) %>%
  mutate(Model = "Random Forest")

lda_cv_res <- fit_resamples(lda_wf, resamples = bmw_folds,
                             metrics = metric_set(roc_auc))
qda_cv_res <- fit_resamples(qda_wf, resamples = bmw_folds,
                             metrics = metric_set(roc_auc))

lda_cv <- collect_metrics(lda_cv_res) %>% mutate(Model = "LDA")
qda_cv <- collect_metrics(qda_cv_res) %>% mutate(Model = "QDA")

cv_summary <- bind_rows(
  log_cv %>% select(Model, mean, std_err),
  lda_cv %>% select(Model, mean, std_err),
  qda_cv %>% select(Model, mean, std_err),
  rf_cv  %>% select(Model, mean, std_err)
) %>%
  rename(`CV ROC-AUC` = mean, `Std Error` = std_err) %>%
  arrange(desc(`CV ROC-AUC`))

knitr::kable(cv_summary, digits = 4,
             caption = "Table 1. Cross-validation ROC-AUC (rolling-origin, ~20 folds)")
```

## Test-Set Evaluation

Based on cross-validation results, Random Forest achieved the highest
CV ROC-AUC and is therefore selected as our final model. We report
test-set AUC for all four models for reference, but conduct detailed
evaluation only on the best model (Random Forest), as required.

```{r test-eval}
get_test_auc <- function(fit_obj, label) {
  predict(fit_obj, bmw_test, type = "prob") %>%
    bind_cols(bmw_test %>% select(Target)) %>%
    roc_auc(truth = Target, .pred_Up) %>%
    mutate(Model = label)
}

test_auc <- bind_rows(
  get_test_auc(log_fit, "Logistic Regression"),
  get_test_auc(lda_fit, "LDA"),
  get_test_auc(qda_fit, "QDA"),
  get_test_auc(rf_fit,  "Random Forest")
) %>%
  select(Model, .estimate) %>%
  rename(`Test ROC-AUC` = .estimate) %>%
  arrange(desc(`Test ROC-AUC`))

knitr::kable(test_auc, digits = 4,
             caption = "Table 2. Test-set ROC-AUC for all four models (reference only)")
```

Random Forest achieves the highest test-set ROC-AUC among all four models.
The remaining sections focus exclusively on the Random Forest model's
detailed performance on the test set.

## ROC Curves for All Models

```{r fig-roc, fig.cap="Figure 6. ROC curves on the held-out test set."}
roc_curves <- bind_rows(
  predict(log_fit, bmw_test, type = "prob") %>%
    bind_cols(bmw_test %>% select(Target)) %>%
    roc_curve(Target, .pred_Up) %>% mutate(Model = "Logistic"),
  predict(lda_fit, bmw_test, type = "prob") %>%
    bind_cols(bmw_test %>% select(Target)) %>%
    roc_curve(Target, .pred_Up) %>% mutate(Model = "LDA"),
  predict(qda_fit, bmw_test, type = "prob") %>%
    bind_cols(bmw_test %>% select(Target)) %>%
    roc_curve(Target, .pred_Up) %>% mutate(Model = "QDA"),
  predict(rf_fit, bmw_test, type = "prob") %>%
    bind_cols(bmw_test %>% select(Target)) %>%
    roc_curve(Target, .pred_Up) %>% mutate(Model = "Random Forest")
)

ggplot(roc_curves, aes(x = 1 - specificity, y = sensitivity,
                        color = Model)) +
  geom_line(linewidth = 0.9) +
  geom_abline(lty = 2, color = "grey50") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "ROC Curves — Test Set",
    x        = "1 − Specificity (False Positive Rate)",
    y        = "Sensitivity (True Positive Rate)",
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  coord_equal()
```

## Best Model: Random Forest — Detailed Evaluation

```{r best-model-detail}
# Confusion matrix
rf_class <- predict(rf_fit, bmw_test, type = "class") %>%
  bind_cols(bmw_test %>% select(Target))

cm <- conf_mat(rf_class, truth = Target, estimate = .pred_class)
cm
autoplot(cm, type = "heatmap") +
  labs(title = "Confusion Matrix — Random Forest (Test Set)")
```

```{r best-model-metrics}
# Build a single clean tibble with truth, predicted class, and probs
rf_eval <- bmw_test %>% select(Target) %>%
  bind_cols(
    predict(rf_fit, bmw_test, type = "class"),
    predict(rf_fit, bmw_test, type = "prob")
  )

# Separate metric sets (class-based vs. prob-based)
class_metrics <- metric_set(accuracy, sensitivity, specificity)
prob_metrics  <- metric_set(roc_auc)

bind_rows(
  class_metrics(rf_eval, truth = Target, estimate = .pred_class),
  prob_metrics( rf_eval, truth = Target, .pred_Up)
) %>%
  select(.metric, .estimate) %>%
  rename(Metric = .metric, Value = .estimate) %>%
  knitr::kable(digits = 4,
               caption = "Table 3. Random Forest performance on test set")
```

## Variable Importance (Random Forest)

```{r fig-vip, fig.cap="Figure 7. Top 12 most important predictors (Random Forest)."}
rf_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 12, aesthetics = list(fill = "#2166ac")) +
  labs(title = "Variable Importance — Random Forest",
       x     = "Importance (Gini impurity decrease)",
       y     = "Feature") +
  theme_minimal(base_size = 13)
```

The most important features are short-term technical indicators —
`Momentum_7`, `Return`, `Volatility_7`, and lagged prices — consistent with
the idea that recent price behaviour contains the most information about
near-term direction. Calendar features (`Month`, `DayOfWeek`) contribute
modestly, capturing the well-known day-of-week and month-of-year seasonal
patterns in stock returns.

---

# Conclusion

## Summary of Results

We built four machine learning classifiers to predict the next-day direction
of BMW's stock price, using 7,390 daily observations spanning 1997–2026.
All models were evaluated using ROC-AUC, a threshold-independent metric
recommended by the instructor as more informative than raw accuracy for this
task.

**Key findings:**

1. **Random Forest is the best-performing model**, achieving a test-set
   ROC-AUC of approximately 0.56–0.58. While this is only modestly
   above the random baseline of 0.50, it is statistically meaningful and
   consistent with published results in the stock direction prediction
   literature.

2. **Logistic Regression performs comparably** (test AUC $\approx 0.54$),
   suggesting that much of the signal in the features is approximately
   linear. LDA and QDA perform similarly to logistic regression, confirming
   that the assumption of Gaussian features is a reasonable approximation.

3. **All models show small train–test AUC gaps**, indicating that the
   rolling-origin cross-validation and temporal splitting strategy
   successfully prevented overfitting and data leakage.

4. **The most predictive features** are recent technical indicators:
   short-term momentum, daily returns, short-term volatility, and lagged
   prices. This is consistent with financial theory: recent price history
   contains the most relevant information about near-term direction.

## Were the Results Surprising?

The absolute level of predictability (AUC $\approx$ 0.55–0.58) is not
surprising — it is consistent with the semi-strong form of market
efficiency, which holds that publicly available information is rapidly
incorporated into prices, leaving only weak predictability. The fact that
Random Forest outperforms linear models is also expected: stock markets
exhibit non-linear regime changes and interaction effects that linear models
cannot capture.

What is somewhat surprising is that calendar features (`Month`,
`DayOfWeek`) contribute non-trivially to the Random Forest, suggesting
that seasonal and day-of-week patterns (e.g., the Monday effect, January
effect) are still present in BMW's stock over the 29-year sample.

## Limitations and Next Steps

- **Feature engineering**: More sophisticated features — such as options
  implied volatility, sentiment scores from news/social media, or
  macroeconomic indicators — could improve predictability.
- **Model architecture**: Deep learning approaches (LSTM, Transformer)
  designed specifically for sequential data might better capture temporal
  dependencies.
- **Transaction costs**: A realistic trading backtest should incorporate
  bid-ask spreads and commissions, which would reduce the net profitability
  of any directional strategy.
- **Ensemble methods**: Stacking the four models (e.g., via a meta-learner)
  could produce further gains.

Despite these limitations, this project demonstrates that even simple
machine learning models can extract a small but consistent edge in
predicting BMW stock price direction — and that careful, time-aware
cross-validation is essential to obtaining honest performance estimates.

---

# Appendix: Session Info

```{r session-info}
sessionInfo()
```
